\documentclass[manuscript,nonacm]{acmart}

\usepackage{array}
\usepackage{mathtools}
\usepackage{xspace}
\usepackage{stmaryrd}
\usepackage{mathpartir}

\setcopyright{none}
\copyrightyear{}
\acmYear{}
\acmDOI{}
\acmConference{}{}{}
\acmISBN{}

\newcommand{\lang}{\textsc{SwiftyPOPL}}
\newcommand{\kw}[1]{\ensuremath{\mathsf{#1}}}
\newcommand{\Int}{\kw{Int}}
\newcommand{\Bool}{\kw{Bool}}
\newcommand{\String}{\kw{String}}
\newcommand{\Unit}{\kw{Unit}}
\newcommand{\Nat}{\kw{Nat}}
\newcommand{\Self}{\kw{Self}\xspace}
\newcommand{\Any}[1]{\kw{any}\ #1}
\newcommand{\ok}{\ \kw{ok}}
\newcommand{\dom}{\operatorname{dom}}
\newcommand{\fv}{\operatorname{fv}}
\newcommand{\drop}{\operatorname{drop}_1}
\newcommand{\subst}[3]{[#2\mapsto #1]#3}
\newcommand{\ftype}[3]{\forall #1.\ (#2) \to #3}
\newcommand{\emptyctx}{\cdot}
\newcommand{\sctx}{\Psi;\Omega;\Gamma;\rho}
\newcommand{\ectx}{\Psi;\Omega;\Gamma;\rho;\Delta}
\newcommand{\has}{\vdash}
\newcommand{\modelsP}{\mathrel{\triangleright}}
\newcommand{\elab}{\leadsto}
\newcommand{\trans}[1]{\llbracket #1 \rrbracket}
\newcommand{\transSelf}[2]{\llbracket #2 \rrbracket^{#1}}
\newcommand{\lambdastar}{\lambda^{\star}}
\newcommand{\appstar}{\mathsf{app}^{\star}}
\newcommand{\arrstar}{\mathsf{arr}^{\star}}
\newcommand{\Dict}{\mathsf{Dict}}
\newcommand{\pack}{\kw{pack}}
\newcommand{\unpack}{\kw{unpack}}
\newcommand{\tin}{\kw{in}}

\begin{document}

\title{\lang: A Swift-like Surface Language with Protocol Elaboration}

\author{Runbang Yan}
\email{rbyan25@stu.pku.edu.cn}
\affiliation{%
  \institution{School of Computer Science, Peking University}
  \city{Beijing}
  \country{China}
}

\renewcommand{\shortauthors}{Yan}

\begin{abstract}
\lang\ is a small Swift-like surface language designed to expose the static structure behind Swift protocols, generic constraints, and existential values.
The implementation parses and checks a nominal surface language, then elaborates it into a core calculus with System F, records, and existential types.
This report presents the motivation, surface syntax, surface type system, elaboration rules, and a paper proof that elaboration preserves typing.
We intentionally rely on the standard soundness theorem for the core calculus and focus on the safety of the translation from the surface language to that core.
\end{abstract}

\maketitle

\section{Motivation}

Swift programmers write nominal declarations such as structs, protocols, extensions, generic functions, and values of type \Any{P}.
The operational meaning of these constructs is less direct than their syntax suggests.
Protocol requirements are implemented by methods on concrete structs, generic constraints need runtime evidence, and existential protocol values hide the concrete type of their payload.
\lang\ makes this structure explicit by elaborating surface programs into a small core language.

The design goal is not to implement all of Swift.
Instead, the language isolates a proof-relevant subset:
nominal structs, protocol requirements, extension-based conformances, constrained generics, Swift-style argument labels, and existential protocol packages.
The central metatheoretic claim is elaboration preservation:
if a surface program type checks, the elaborated core program has the translated type.
Since core soundness is standard, this is the main nontrivial proof obligation for the surface language.

\section{Surface Syntax Design}

We write $S$ for struct names, $P,Q$ for protocol names, $X,Y$ for type variables, $x,f,m$ for value variables and function or method names, and $\ell$ for external argument labels.
The distinguished label $\epsilon$ denotes an unlabeled positional parameter.

\[
\begin{array}{rcll}
\tau & ::= & \Int \mid \Bool \mid \String \mid \Unit
  & \text{base types} \\
& \mid & S
  & \text{nominal struct type} \\
& \mid & X
  & \text{type variable} \\
& \mid & \Self
  & \text{contextual self type} \\
& \mid & \Any{P}
  & \text{existential protocol type} \\
& \mid & \ftype{\overline{X : \overline{P}}}{\overline{\ell : \tau}}{\tau}
  & \text{function scheme} \\[1mm]

e & ::= & n \mid b \mid s \mid x \mid \kw{self}
  & \text{literals and variables} \\
& \mid & {!}e \mid e\mathbin{\&\&}e \mid e\mathbin{||}e
  & \text{boolean operators} \\
& \mid & e+e \mid e-e \mid e<e \mid e>e \mid e\leq e \mid e\geq e
  & \text{integer operators} \\
& \mid & \kw{func}\langle\overline{X : \overline{P}}\rangle
          (\overline{\ell\ x : \tau}) \to \tau\ \{ ss \}
  & \text{function value} \\
& \mid & e(\overline{a}) \mid e\langle\overline{\tau}\rangle
  & \text{call and type application} \\
& \mid & e.m \mid S(\overline{\ell : e})
  & \text{member access and construction} \\
& \mid & \kw{if}\ e\ \{ ss \}\ \kw{else}\ \{ ss \}
  & \text{conditional expression} \\
& \mid & e\ \kw{as}\ \Any{P}
  & \text{existential package} \\[1mm]

a & ::= & e \mid \ell : e
  & \text{positional or labeled argument} \\[1mm]

ss & ::= & \epsilon \mid stmt;\ ss
  & \text{statement sequence} \\
stmt & ::= & e \mid d
  & \text{expression or declaration} \\[1mm]

d & ::= & \kw{let}\ x : \tau = e
  & \text{value declaration} \\
& \mid & \kw{func}\ f\langle\overline{X : \overline{P}}\rangle
          (\overline{\ell\ x : \tau}) \to \tau\ \{ ss \}
  & \text{function definition} \\
& \mid & \kw{struct}\ S\ \{ \overline{\kw{let}\ x : \tau};\ \overline{md} \}
  & \text{struct definition} \\
& \mid & \kw{protocol}\ P\ \{ \overline{msig} \}
  & \text{protocol definition} \\
& \mid & \kw{extension}\ S : P\ \{ \overline{md} \}
  & \text{conformance definition}
\end{array}
\]

Method definitions and protocol requirements have the same shape:
\[
\begin{array}{rcl}
md & ::= &
\kw{func}\ m\langle\overline{X : \overline{P}}\rangle
(\overline{\ell\ x : \tau}) \to \tau\ \{ ss \}
\\
msig & ::= &
\kw{func}\ m\langle\overline{X : \overline{P}}\rangle
(\overline{\ell : \tau}) \to \tau .
\end{array}
\]

The current implementation fragment maps surface $\Int$ to core $\Nat$.
Therefore subtraction is saturating: after elaboration, $0-1$ evaluates to $0$ because core $\kw{pred}\ 0$ evaluates to $0$.
String concatenation is intentionally absent because the current core language has no string-concatenation primitive.
For function parameters, the parser follows Swift's label convention: when no external label is explicitly provided, the external label defaults to the parameter name; writing $\_$ makes the parameter unlabeled.
The surface fragment formalized in this report does not support recursive user-defined functions.
The core language still contains $\kw{fix}$ because the elaboration of built-in arithmetic operators uses it internally.

\section{Surface Type System}

\subsection{Environments and Method Schemes}

The type system uses a global environment $\Psi$, a generic constraint environment $\Omega$, a local value environment $\Gamma$, and a \Self resolver $\rho$.

\[
\begin{array}{rcll}
\Psi & ::= & (\Psi_S,\Psi_P,\Psi_C,\Psi_M,\Psi_V)
  & \text{global environment} \\
\Psi_S(S) &=& \{\overline{x:\tau}\}
  & \text{struct fields} \\
\Psi_P(P) &=& \{\overline{m:\sigma}\}
  & \text{protocol requirements} \\
\Psi_C(S,P) &=& \{\overline{m:\sigma}\}
  & \text{extension methods implementing } S:P \\
\Psi_M(S) &=& \{\overline{m:\sigma}\}
  & \text{struct and extension methods of } S \\
\Psi_V(x) &=& \sigma
  & \text{top-level value or function type} \\
\Omega & ::= & \emptyctx \mid \Omega,X:\overline{P}
  & \text{type-variable constraints} \\
\Gamma & ::= & \emptyctx \mid \Gamma,x:\tau
  & \text{local values} \\
\rho & ::= & \bot \mid S \mid P
  & \text{\Self resolver.}
\end{array}
\]

A method scheme always has an explicit receiver parameter.
For a method of struct $S$, the receiver has type $S$.
For a requirement in protocol $P$, the receiver has type \Self under resolver $P$.
This convention is what lets member access elaborate into ordinary function application.

\subsection{Well-Formed Types and \Self}

The judgment $\Psi;\Omega;\rho \has \tau\ok$ means that $\tau$ is well formed.
Base types are always well formed; nominal types must be declared; type variables must be in $\Omega$; $\Self$ is available only when $\rho\neq\bot$; $\Any{P}$ is well formed only when $P$ is declared; and function schemes are checked under their own type parameters and declared protocol constraints.

\begin{mathpar}
\inferrule*[Right=\textsc{WF-Base}]
  { \tau_b\in\{\Int,\Bool,\String,\Unit\} }
  { \Psi;\Omega;\rho \has \tau_b\ok }

\inferrule*[Right=\textsc{WF-Struct}]
  { S\in\dom(\Psi_S) }
  { \Psi;\Omega;\rho \has S\ok }

\inferrule*[Right=\textsc{WF-TVar}]
  { X\in\dom(\Omega) }
  { \Psi;\Omega;\rho \has X\ok }
\end{mathpar}

\begin{mathpar}
\inferrule*[Right=\textsc{WF-Self}]
  { \rho\neq\bot }
  { \Psi;\Omega;\rho \has \Self\ok }

\inferrule*[Right=\textsc{WF-Any}]
  { P\in\dom(\Psi_P) }
  { \Psi;\Omega;\rho \has \Any{P}\ok }
\end{mathpar}

\begin{mathpar}
\inferrule*[Right=\textsc{WF-Fun}]
  {
    \overline{X_i}\ \text{distinct}
    \\
    \forall i,k.\ P_{ik}\in\dom(\Psi_P)
    \\
    \Psi;\Omega,\overline{X_i:\overline{P_{ik}}};\rho
      \has \overline{\tau_j}\ok
    \\
    \Psi;\Omega,\overline{X_i:\overline{P_{ik}}};\rho
      \has \tau_r\ok
  }
  {
    \Psi;\Omega;\rho
      \has
      \ftype{\overline{X_i:\overline{P_{ik}}}}
            {\overline{\ell_j:\tau_j}}
            {\tau_r}
      \ok
  }
\end{mathpar}

Inside a struct or extension declaration, \Self resolves to the concrete receiver struct.
Inside a protocol declaration, \Self remains abstract and can appear in requirement signatures.

\subsection{Protocol Satisfaction}

The judgment $\Psi;\Omega \has \tau \modelsP P$ means that $\tau$ is known to satisfy protocol $P$.

\begin{mathpar}
\inferrule*[Right=\textsc{Sat-Struct}]
  { (S,P)\in\dom(\Psi_C) }
  { \Psi;\Omega \has S \modelsP P }

\inferrule*[Right=\textsc{Sat-TVar}]
  { X:\overline{P}\in\Omega \\ P\in\overline{P} }
  { \Psi;\Omega \has X \modelsP P }
\end{mathpar}

Only concrete structs and constrained type variables satisfy protocols.
In particular, $\Any{P}$ does not itself satisfy $P$.
An existential package must be opened by elaboration before its dictionary can be used.

\subsection{Expression Typing}

The main expression judgment is
\[
  \Psi;\Omega;\Gamma;\rho \has e:\tau.
\]
The following rules cover the expression forms used by the preservation proof.

\begin{mathpar}
\inferrule*[Right=\textsc{T-Int}]
  { }
  { \Psi;\Omega;\Gamma;\rho \has n:\Int }

\inferrule*[Right=\textsc{T-Bool}]
  { }
  { \Psi;\Omega;\Gamma;\rho \has b:\Bool }

\inferrule*[Right=\textsc{T-String}]
  { }
  { \Psi;\Omega;\Gamma;\rho \has s:\String }

\inferrule*[Right=\textsc{T-Var}]
  { x:\tau\in\Gamma \text{ or } x:\tau\in\Psi_V }
  { \Psi;\Omega;\Gamma;\rho \has x:\tau }

\inferrule*[Right=\textsc{T-Self}]
  { \rho=S }
  { \Psi;\Omega;\Gamma;\rho \has \kw{self}:S }

\inferrule*[Right=\textsc{T-Not}]
  { \Psi;\Omega;\Gamma;\rho \has e:\Bool }
  { \Psi;\Omega;\Gamma;\rho \has {!}e:\Bool }

\inferrule*[Right=\textsc{T-IntArith}]
  {
    \oplus\in\{+,-\} \\
    \Psi;\Omega;\Gamma;\rho \has e_1:\Int \\
    \Psi;\Omega;\Gamma;\rho \has e_2:\Int
  }
  { \Psi;\Omega;\Gamma;\rho \has e_1\oplus e_2:\Int }

\inferrule*[Right=\textsc{T-IntCmp}]
  {
    \bowtie\in\{<,>,\leq,\geq\} \\
    \Psi;\Omega;\Gamma;\rho \has e_1:\Int \\
    \Psi;\Omega;\Gamma;\rho \has e_2:\Int
  }
  { \Psi;\Omega;\Gamma;\rho \has e_1\bowtie e_2:\Bool }

\inferrule*[Right=\textsc{T-BoolBin}]
  {
    \otimes\in\{\mathbin{\&\&},\mathbin{||}\} \\
    \Psi;\Omega;\Gamma;\rho \has e_1:\Bool \\
    \Psi;\Omega;\Gamma;\rho \has e_2:\Bool
  }
  { \Psi;\Omega;\Gamma;\rho \has e_1\otimes e_2:\Bool }
\end{mathpar}

Function values, calls, and type applications are typed using function schemes.
Generic constraints are checked at type application time by requiring protocol satisfaction evidence for the chosen type arguments.

\begin{mathpar}
\inferrule*[Right=\textsc{T-Fun}]
  {
    \Omega'=\Omega,\overline{X_i:\overline{P_i}} \\
    \Gamma'=\Gamma,\overline{x_j:\tau_j} \\
    \Psi;\Omega';\Gamma';\rho \has ss:\tau_r
  }
  {
    \Psi;\Omega;\Gamma;\rho \has
    \kw{func}\langle\overline{X_i:\overline{P_i}}\rangle
      (\overline{\ell_j\ x_j:\tau_j}) \to \tau_r\ \{ss\}
    :
    \ftype{\overline{X_i:\overline{P_i}}}
           {\overline{\ell_j:\tau_j}}
           {\tau_r}
  }

\inferrule*[Right=\textsc{T-TypeApp}]
  {
    \Psi;\Omega;\Gamma;\rho \has e:
      \ftype{\overline{X_i:\overline{P_i}}}
             {\overline{\ell_j:\tau_j}}
             {\tau_r} \\
    \overline{\Psi;\Omega \has \tau_i' \modelsP P
      \quad\text{for every }P\in\overline{P_i}}
  }
  {
    \Psi;\Omega;\Gamma;\rho \has
      e\langle\overline{\tau_i'}\rangle :
      \subst{\overline{\tau_i'}}{\overline{X_i}}
      \left(
        \ftype{\emptyctx}{\overline{\ell_j:\tau_j}}{\tau_r}
      \right)
  }

\inferrule*[Right=\textsc{T-Call}]
  {
    \Psi;\Omega;\Gamma;\rho \has e_f:
      \ftype{\emptyctx}{\overline{\ell_i:\tau_i}}{\tau_r} \\
    \Psi;\Omega;\Gamma;\rho
      \has \overline{a}\Leftarrow\overline{\ell_i:\tau_i}
  }
  { \Psi;\Omega;\Gamma;\rho \has e_f(\overline{a}):\tau_r }
\end{mathpar}

The auxiliary argument judgment checks positional and labeled arguments against formal parameters after deterministic label matching.
This keeps the call rule independent of the implementation details of Swift-style argument-label ordering.

\begin{mathpar}
\inferrule*[Right=\textsc{T-Construct}]
  {
    \Psi_S(S)=\{\overline{x_i:\tau_i}\} \\
    \Psi;\Omega;\Gamma;\rho
      \has \overline{\ell_i:e_i}\Leftarrow\overline{x_i:\tau_i}
  }
  { \Psi;\Omega;\Gamma;\rho \has S(\overline{\ell_i:e_i}):S }

\inferrule*[Right=\textsc{T-Field}]
  {
    \Psi;\Omega;\Gamma;\rho \has e:S \\
    \Psi_S(S)(x)=\tau
  }
  { \Psi;\Omega;\Gamma;\rho \has e.x:\tau }

\inferrule*[Right=\textsc{T-StructMethod}]
  {
    \Psi;\Omega;\Gamma;\rho \has e:S \\
    \Psi_M(S)(m)=
      \ftype{\overline{X_i:\overline{P_i}}}
             {\epsilon:S,\overline{\ell_j:\tau_j}}
             {\tau_r}
  }
  {
    \Psi;\Omega;\Gamma;\rho \has e.m:
      \ftype{\overline{X_i:\overline{P_i}}}
             {\overline{\ell_j:\tau_j}}
             {\tau_r}
  }

\inferrule*[Right=\textsc{T-ConstrainedMethod}]
  {
    \Psi;\Omega;\Gamma;\rho \has e:X \\
    X:\overline{P}\in\Omega \\
    P_k\in\overline{P} \\
    \Psi_P(P_k)(m)=
      \ftype{\overline{Y_i:\overline{Q_i}}}
             {\epsilon:\Self,\overline{\ell_j:\tau_j}}
             {\tau_r}
  }
  {
    \Psi;\Omega;\Gamma;\rho \has e.m:
      [\Self\mapsto X]
      \left(
        \ftype{\overline{Y_i:\overline{Q_i}}}
               {\overline{\ell_j:\tau_j}}
               {\tau_r}
      \right)
  }
\end{mathpar}

Existential method selection is the controlled surface elimination form for $\Any{P}$.
It is allowed only when the visible method signature does not mention protocol \Self.
This prevents the hidden witness type from escaping.
Generic existential methods keep their method-local type parameters; only the hidden receiver is supplied by opening the package.

\begin{mathpar}
\inferrule*[Right=\textsc{T-ExistentialMethod}]
  {
    \Psi;\Omega;\Gamma;\rho \has e:\Any{P} \\
    \Psi_P(P)(m)=
      \ftype{\overline{Y_i:\overline{Q_i}}}
             {\epsilon:\Self,\overline{\ell_j:\tau_j}}
             {\tau_r} \\
    \Self\notin\fv(\overline{\tau_j}) \\
    \Self\notin\fv(\tau_r)
  }
  {
    \Psi;\Omega;\Gamma;\rho \has e.m:
      \ftype{\overline{Y_i:\overline{Q_i}}}
             {\overline{\ell_j:\tau_j}}
             {\tau_r}
  }

\inferrule*[Right=\textsc{T-IfElse}]
  {
    \Psi;\Omega;\Gamma;\rho \has e_c:\Bool \\
    \Psi;\Omega;\Gamma;\rho \has ss_1:\tau \\
    \Psi;\Omega;\Gamma;\rho \has ss_2:\tau
  }
  {
    \Psi;\Omega;\Gamma;\rho \has
      \kw{if}\ e_c\ \{ss_1\}\ \kw{else}\ \{ss_2\}:\tau
  }

\inferrule*[Right=\textsc{T-AsAny}]
  {
    \Psi;\Omega;\Gamma;\rho \has e:\tau \\
    \Psi;\Omega \has \tau \modelsP P
  }
  { \Psi;\Omega;\Gamma;\rho \has e\ \kw{as}\ \Any{P}:\Any{P} }
\end{mathpar}

\subsection{Declarations and Programs}

Top-level checking is collection based.
First, all top-level declarations are collected into $\Psi$.
Collection requires distinct top-level names, declared nominal references, and no duplicate methods within a single owner.
Then each declaration is checked against $\Psi$.
A conformance declaration $\kw{extension}\ S:P$ must implement every requirement of $P$, and each implementation type must match the requirement after substituting $[\Self\mapsto S]$.

\section{Elaboration to Core}

\subsection{Core Target Language}

The target calculus is System F with records, existentials, conditionals, recursion, and natural-number primitives.
We rely on the standard type soundness theorem for this core.

\[
\begin{array}{rcll}
A & ::= & \Nat \mid \Bool \mid \String \mid \Unit
  & \text{base types} \\
& \mid & X \mid A\to A \mid \forall X.A
  & \text{variables, functions, universals} \\
& \mid & \{\overline{l:A}\} \mid \exists X.A
  & \text{records and existentials} \\[1mm]
t & ::= & n \mid b \mid s \mid () \mid x
  & \text{literals and variables} \\
& \mid & \lambda x:A.t \mid t\ t \mid \Lambda X.t \mid t[A]
  & \text{abstraction and application} \\
& \mid & \{\overline{l=t}\} \mid t.l
  & \text{records} \\
& \mid & \kw{if}\ t\ \kw{then}\ t\ \kw{else}\ t
  & \text{conditionals} \\
& \mid & \kw{fix}\ t \mid \kw{succ}\ t \mid \kw{pred}\ t \mid \kw{iszero}\ t
  & \text{recursion and naturals} \\
& \mid & \pack[A,t]\ \kw{as}\ \exists X.B
  & \text{existential introduction} \\
& \mid & \unpack\ \{X,x\}=t\ \tin\ t
  & \text{existential elimination.}
\end{array}
\]

The elaboration judgment is
\[
  \Psi;\Omega;\Gamma;\rho;\Delta \has e:\tau \elab t.
\]
The evidence environment $\Delta$ maps constrained type variables to dictionary variables:
$\Delta(X,P)=d_{X,P}$.
Concrete conformances elaborate to top-level dictionary names $\delta(S,P)$; method definitions elaborate to top-level function names $\mu(S,m)$.

\subsection{Type and Dictionary Translation}

Surface types translate to core types as follows.
\[
\begin{array}{rcl}
\trans{\Int} &=& \Nat \\
\trans{\Bool} &=& \Bool \\
\trans{\String} &=& \String \\
\trans{\Unit} &=& \Unit \\
\trans{X} &=& X \\
\trans{S} &=& \{\overline{x:\trans{\tau}}\}
  \quad\text{where }\Psi_S(S)=\{\overline{x:\tau}\} \\
\trans{\Any{P}} &=&
  \exists X.\{\kw{value}:X,\kw{dict}:\Dict(P,X)\} \\
\trans{\ftype{\overline{X_i:\overline{P_i}}}
              {\overline{\ell_j:\tau_j}}
              {\tau_r}}
  &=&
  \forall\overline{X_i}.\
  \trans{\overline{X_i:\overline{P_i}}}
  \to
  \arrstar(\overline{\trans{\tau_j}},\trans{\tau_r}).
\end{array}
\]

The implementation represents a surface zero-argument function as a core function from \Unit.
We write this convention with three meta-level shorthands:
\[
\begin{array}{rclcrcl}
\arrstar(\emptyctx,A) &=& \Unit\to A,
&\quad&
\arrstar(A_1,\ldots,A_n,A) &=& A_1\to\cdots\to A_n\to A \quad(n>0),\\
\lambdastar\ \emptyctx.\ t &=& \lambda\_:\Unit.\ t,
&\quad&
\lambdastar\ \overline{y_i:A_i}.\ t &=& \lambda \overline{y_i:A_i}.\ t \quad(n>0),\\
\appstar(t,\emptyctx) &=& t\ (),
&\quad&
\appstar(t,\overline{u_i}) &=& t\ \overline{u_i} \quad(n>0).
\end{array}
\]
This is an implementation encoding of surface arity; it is not a surface-level identification of \Unit\ arguments with empty argument lists.

For every constraint $X:P$, the translated function type inserts one dictionary argument of type $\Dict(P,X)$ before the ordinary value arguments.
The ordinary translation $\trans{\tau}$ is used after resolution.
Thus \Self that appears in a struct or extension body has already been resolved to the concrete struct type.
Protocol requirements are the only translated types that may still contain an abstract protocol \Self.
For those requirement signatures we use a self-parameterized translation $\transSelf{A}{\tau}$, which is the same as $\trans{\tau}$ except that each occurrence of protocol \Self is translated as the core type $A$.
This corresponds to the implementation's resolved-type substitution of \texttt{AbstractSelfType} before rendering the core type.

A protocol dictionary is a record of methods:
\[
  \Dict(P,A)=
  \{\overline{m_i:\transSelf{A}{\sigma_i}}\}
  \quad\text{where }\Psi_P(P)=\{\overline{m_i:\sigma_i}\}.
\]

\subsection{Evidence Elaboration}

The judgment
\[
  \Psi;\Omega;\Delta \has \tau \modelsP P \elab d
\]
produces the dictionary term witnessing $\tau:P$.

\begin{mathpar}
\inferrule*[Right=\textsc{Ev-Struct}]
  { (S,P)\in\dom(\Psi_C) }
  { \Psi;\Omega;\Delta \has S \modelsP P \elab \delta(S,P) }

\inferrule*[Right=\textsc{Ev-TVar}]
  { X:\overline{P}\in\Omega \\ P\in\overline{P} \\ \Delta(X,P)=d }
  { \Psi;\Omega;\Delta \has X \modelsP P \elab d }
\end{mathpar}

\subsection{Operator Elaboration}

Surface operators elaborate into closed core terms.
{\scriptsize
\[
\begin{array}{rcl}
\mathsf{add}
&\triangleq&
\kw{fix}\,
  (\lambda \mathit{add}:\Nat\to\Nat\to\Nat.\,
   \lambda x:\Nat.\lambda y:\Nat.\,
   \kw{if}\ \kw{iszero}\ x\ \kw{then}\ y\
   \kw{else}\ \kw{succ}\ ((\mathit{add}\ (\kw{pred}\ x))\ y))
\\[1mm]
\mathsf{sub}
&\triangleq&
\kw{fix}\,
  (\lambda \mathit{sub}:\Nat\to\Nat\to\Nat.\,
   \lambda x:\Nat.\lambda y:\Nat.\,
   \kw{if}\ \kw{iszero}\ y\ \kw{then}\ x\
   \kw{else}\ ((\mathit{sub}\ (\kw{pred}\ x))\ (\kw{pred}\ y)))
\\[1mm]
\mathsf{lt}
&\triangleq&
\kw{fix}\,
  (\lambda \mathit{lt}:\Nat\to\Nat\to\Bool.\,
   \lambda x:\Nat.\lambda y:\Nat.\,
   \kw{if}\ \kw{iszero}\ y\ \kw{then}\ \kw{false}\
   \kw{else}\ \kw{if}\ \kw{iszero}\ x\ \kw{then}\ \kw{true}\
   \kw{else}\ ((\mathit{lt}\ (\kw{pred}\ x))\ (\kw{pred}\ y))).
\end{array}
\]
}
These abbreviations have core types
\[
  \has_c \mathsf{add}:\Nat\to\Nat\to\Nat
  \qquad
  \has_c \mathsf{sub}:\Nat\to\Nat\to\Nat
  \qquad
  \has_c \mathsf{lt}:\Nat\to\Nat\to\Bool.
\]

If $e_i\elab t_i$, then the operator cases are:
\[
\begin{array}{rclcrcl}
e_1+e_2 &\elab& \mathsf{add}\ t_1\ t_2
&\quad&
e_1-e_2 &\elab& \mathsf{sub}\ t_1\ t_2 \\
e_1<e_2 &\elab& \mathsf{lt}\ t_1\ t_2
&&
e_1>e_2 &\elab& \mathsf{lt}\ t_2\ t_1 \\
e_1\leq e_2 &\elab&
\kw{if}\ (\mathsf{lt}\ t_2\ t_1)\ \kw{then}\ \kw{false}\ \kw{else}\ \kw{true}
&&
e_1\geq e_2 &\elab&
\kw{if}\ (\mathsf{lt}\ t_1\ t_2)\ \kw{then}\ \kw{false}\ \kw{else}\ \kw{true} \\
{!}e &\elab&
\kw{if}\ t\ \kw{then}\ \kw{false}\ \kw{else}\ \kw{true}
&&
e_1\mathbin{\&\&}e_2 &\elab&
\kw{if}\ t_1\ \kw{then}\ t_2\ \kw{else}\ \kw{false} \\
e_1\mathbin{||}e_2 &\elab&
\kw{if}\ t_1\ \kw{then}\ \kw{true}\ \kw{else}\ t_2.
\end{array}
\]
The boolean connectives preserve short-circuiting because the second operand appears only in the selected branch of a core conditional.

\subsection{Expression Elaboration}

Most expression rules are homomorphic.
The essential non-homomorphic rules are generic application, method lookup, existential packaging, and existential-safe method selection.
The homomorphic rules below translate literals, variables, calls, field access, construction, and conditionals structurally to the corresponding core form.
Primitive operators use the translations from the previous subsection.
These rules do not create new type abstractions, dictionaries, or existential packages.

\begin{mathpar}
\inferrule*[Right=\textsc{E-Int}]
  { }
  { \ectx \has n:\Int \elab n }

\inferrule*[Right=\textsc{E-Bool}]
  { }
  { \ectx \has b:\Bool \elab b }

\inferrule*[Right=\textsc{E-String}]
  { }
  { \ectx \has s:\String \elab s }

\inferrule*[Right=\textsc{E-Var}]
  { x:\tau\in\Gamma\ \text{or}\ \Psi_V(x)=\tau }
  { \ectx \has x:\tau \elab x }

\inferrule*[Right=\textsc{E-Self}]
  { \kw{self}:S\in\Gamma }
  { \ectx \has \kw{self}:S \elab \kw{self} }
\end{mathpar}

\begin{mathpar}
\inferrule*[Right=\textsc{E-Call}]
  {
    \ectx \has e:\ftype{\emptyctx}{\overline{\ell_i:\tau_i}}{\tau_r}
      \elab t \\
    \ectx \has \overline{a_i}\Leftarrow\overline{\ell_i:\tau_i}
      \elab \overline{u_i}
  }
  {
    \ectx \has e(\overline{a_i}):\tau_r
      \elab \appstar(t,\overline{u_i})
  }

\inferrule*[Right=\textsc{E-Construct}]
  {
    \Psi_S(S)=\{\overline{x_i:\tau_i}\} \\
    \ectx \has \overline{a_i}\Leftarrow\overline{x_i:\tau_i}
      \elab \overline{t_i}
  }
  {
    \ectx \has S(\overline{a_i}):S
      \elab \{\overline{x_i=t_i}\}
  }

\inferrule*[Right=\textsc{E-Field}]
  {
    \ectx \has e:S \elab t \\
    \Psi_S(S)(x)=\tau
  }
  {
    \ectx \has e.x:\tau \elab t.x
  }
\end{mathpar}

\begin{mathpar}
\inferrule*[Right=\textsc{E-IfElse}]
  {
    \ectx \has e:\Bool \elab t \\
    \ectx \has ss_1:\tau \elab t_1 \\
    \ectx \has ss_2:\tau \elab t_2
  }
  {
    \ectx \has
      \kw{if}\ e\ \{ss_1\}\ \kw{else}\ \{ss_2\}:\tau
      \elab
      \kw{if}\ t\ \kw{then}\ t_1\ \kw{else}\ t_2
  }
\end{mathpar}

Function values are non-homomorphic because constrained type parameters elaborate to explicit dictionary parameters before the ordinary value parameters.

\begin{mathpar}
\inferrule*[Right=\textsc{E-Fun}]
  {
    \Omega'=\Omega,\overline{X_i:\overline{P_i}} \\
    \Delta'=\Delta,\overline{(X_i,P)=\delta(X_i,P)} \\
    \Gamma_f=\Gamma,\overline{x_j:\tau_j} \\
    \Psi;\Omega';\Gamma_f;\rho;\Delta'
      \has ss:\tau_r \elab t_f
  }
  {
    \ectx \has
      \kw{func}\langle\overline{X_i:\overline{P_i}}\rangle
      (\overline{\ell_j\ x_j:\tau_j})\to\tau_r\ \{ss\}
      :
      \ftype{\overline{X_i:\overline{P_i}}}
             {\overline{\ell_j:\tau_j}}
             {\tau_r}
      \elab M_{\lambda}
  }
\end{mathpar}

\[
M_{\lambda} =
  \Lambda \overline{X_i}.\
  \lambda \overline{\delta(X_i,P):\Dict(P,X_i)}.\
  \lambdastar\ \overline{x_j:\trans{\tau_j}}.\ t_f.
\]

\begin{mathpar}
\inferrule*[Right=\textsc{E-TypeApp}]
  {
    \ectx \has e:
      \ftype{\overline{X_i:\overline{P_i}}}
             {\overline{\ell_j:\tau_j}}
             {\tau_r}
      \elab t \\
    \overline{\Psi;\Omega;\Delta \has \tau_i' \modelsP P
      \elab d_{i,P}
      \quad\text{for every }P\in\overline{P_i}}
  }
  {
    \ectx \has e\langle\overline{\tau_i'}\rangle:
      \subst{\overline{\tau_i'}}{\overline{X_i}}
      \left(
        \ftype{\emptyctx}{\overline{\ell_j:\tau_j}}{\tau_r}
      \right)
      \elab
      t[\overline{\trans{\tau_i'}}]\ \overline{d_{i,P}}
  }

\inferrule*[Right=\textsc{E-StructMethod}]
  {
    \ectx \has e:S \elab t \\
    \Psi_M(S)(m)=
      \ftype{\overline{X_i:\overline{P_i}}}
             {\epsilon:S,\overline{\ell_j:\tau_j}}
             {\tau_r}
  }
  {
    \ectx \has e.m:
      \ftype{\overline{X_i:\overline{P_i}}}
             {\overline{\ell_j:\tau_j}}
             {\tau_r}
      \elab M_S
  }

\inferrule*[Right=\textsc{E-ConstrainedMethod}]
  {
    \ectx \has e:X \elab t \\
    X:\overline{P}\in\Omega \\
    P_k\in\overline{P} \\
    \Psi_P(P_k)(m)=
      \ftype{\overline{Y_i:\overline{Q_i}}}
             {\epsilon:\Self,\overline{\ell_j:\tau_j}}
             {\tau_r} \\
    \Delta(X,P_k)=d
  }
  {
    \ectx \has e.m:
      [\Self\mapsto X]
      \left(
        \ftype{\overline{Y_i:\overline{Q_i}}}
               {\overline{\ell_j:\tau_j}}
               {\tau_r}
      \right)
      \elab M_X
  }

\inferrule*[Right=\textsc{E-AsAny}]
  {
    \ectx \has e:\tau \elab t \\
    \Psi;\Omega;\Delta \has \tau \modelsP P \elab d
  }
  {
    \ectx \has e\ \kw{as}\ \Any{P}:\Any{P}
      \elab
      \pack[\trans{\tau},\{\kw{value}=t,\kw{dict}=d\}]
      \ \kw{as}\
      \exists X.\{\kw{value}:X,\kw{dict}:\Dict(P,X)\}
  }
\end{mathpar}

where the method-selection target terms are:
\[
\begin{aligned}
M_S &=
  \Lambda \overline{X_i}.\
  \lambda \overline{\delta(X_i,P):\Dict(P,X_i)}.\
  \lambdastar\ \overline{y_j:\trans{\tau_j}}.\
  \appstar(
    \mu(S,m)[\overline{X_i}]\ \overline{\delta(X_i,P)}\ t,
    \overline{y_j}),\\
M_X &=
  \Lambda \overline{Y_i}.\
  \lambda \overline{\delta(Y_i,Q):\Dict(Q,Y_i)}.\
  \lambdastar\ \overline{y_j:\trans{[\Self\mapsto X]\tau_j}}.\
  \appstar(
    d.m[\overline{Y_i}]\ \overline{\delta(Y_i,Q)}\ t,
    \overline{y_j}).
\end{aligned}
\]

Existential method selection opens the package and immediately calls the dictionary method on the hidden receiver.
The visible function returned by member access abstracts over method type parameters, required dictionaries, and the remaining method arguments.

\begin{mathpar}
\inferrule*[Right=\textsc{E-ExistentialMethod}]
  {
    \ectx \has e:\Any{P}\elab t \\
    \Psi_P(P)(m)=
      \ftype{\overline{Y_i:\overline{Q_i}}}
             {\epsilon:\Self,\overline{\ell_j:\tau_j}}
             {\tau_r} \\
    \Self\notin\fv(\overline{\tau_j}) \\
    \Self\notin\fv(\tau_r)
  }
  {
    \ectx \has e.m:
      \ftype{\overline{Y_i:\overline{Q_i}}}
             {\overline{\ell_j:\tau_j}}
             {\tau_r}
      \elab M_{\exists}
  }
\end{mathpar}

\[
\begin{aligned}
M_{\exists} =
  \Lambda \overline{Y_i}.\
  \lambda \overline{\delta(Y_i,Q):\Dict(Q,Y_i)}.\
  \lambdastar\ \overline{y_j:\trans{\tau_j}}.\
  \unpack\ \{X,z\}=t\ \tin\
  \appstar(
    z.\kw{dict}.m[\overline{Y_i}]\
      \overline{\delta(Y_i,Q)}\
      z.\kw{value},
    \overline{y_j}).
\end{aligned}
\]

Inside the unpack body, $z.\kw{value}:X$ and $z.\kw{dict}:\Dict(P,X)$.
The dictionary method has the requirement type translated with protocol \Self interpreted as $X$.
The side conditions ensure that the visible argument and result types do not mention this local $X$, so the result type of the unpack is allowed to escape.

\subsection{Statement Sequence Elaboration}

Statement sequences elaborate to core expressions.
An empty sequence produces unit; a sequence whose first expression is not the final result evaluates it and discards the value; local declarations elaborate by introducing local bindings.

\begin{mathpar}
\inferrule*[Right=\textsc{ES-Empty}]
  { }
  { \ectx \has \epsilon:\Unit \elab () }

\inferrule*[Right=\textsc{ES-Expr-Last}]
  { \ectx \has e:\tau \elab t }
  { \ectx \has e;\epsilon:\tau \elab t }

\inferrule*[Right=\textsc{ES-Expr-Cons}]
  {
    \ectx \has e:\tau_1 \elab t_1 \\
    \ectx \has ss:\tau_2 \elab t_2
  }
  {
    \ectx \has e;ss:\tau_2
      \elab
      (\lambda \_:\trans{\tau_1}.\ t_2)\ t_1
  }

\inferrule*[Right=\textsc{ES-Let}]
  {
    \ectx \has e:\tau_x \elab t_x \\
    \Psi;\Omega;\Gamma,x:\tau_x;\rho;\Delta \has ss:\tau \elab t
  }
  {
    \ectx \has \kw{let}\ x:\tau_x=e;\ ss:\tau
      \elab
      (\lambda x:\trans{\tau_x}.\ t)\ t_x
  }

\inferrule*[Right=\textsc{ES-Fun}]
  {
    \sigma=
      \ftype{\overline{X_i:\overline{P_i}}}
             {\overline{\ell_j:\tau_j}}
             {\tau_r} \\
    \Omega'=\Omega,\overline{X_i:\overline{P_i}} \\
    \Delta'=\Delta,\overline{(X_i,P)=\delta(X_i,P)} \\
    \Gamma_f=\Gamma,\overline{x_j:\tau_j} \\
    \Psi;\Omega';\Gamma_f;\rho;\Delta' \has ss_f:\tau_r \elab t_f \\
    \Psi;\Omega;\Gamma,f:\sigma;\rho;\Delta \has ss:\tau \elab t
  }
  {
    \ectx \has
      \kw{func}\ f\langle\overline{X_i:\overline{P_i}}\rangle
      (\overline{\ell_j\ x_j:\tau_j})\to\tau_r\ \{ss_f\};\ ss
      :\tau
      \elab
      (\lambda f:\trans{\sigma}.\ t)\ M_f
  }
\end{mathpar}

\[
M_f =
  \Lambda \overline{X_i}.\
  \lambda \overline{\delta(X_i,P):\Dict(P,X_i)}.\
  \lambdastar\ \overline{x_j:\trans{\tau_j}}.\ t_f.
\]

Local function declarations use the same generated function term as \textsc{E-Fun}, then bind it locally before elaborating the rest of the sequence.
The function name is available only in the following sequence, not in its own body.

\subsection{Declarations}

Struct and extension methods elaborate to top-level core functions.
For a method body checked with receiver struct $S$, the generated function abstracts over method type parameters, dictionary parameters for method constraints, the receiver $\kw{self}:S$, and then ordinary parameters.
The declaration elaboration rules are written schematically as follows.
For top-level lets and functions:
\[
\begin{array}{rcl}
\kw{let}\ x:\tau=e &\elab& x=t
  \quad\text{if }\Psi;\emptyctx;\emptyctx;\bot;\emptyctx
  \has e:\tau\elab t,\\[1mm]
\kw{func}\ f:\sigma\ \{ss_f\} &\elab& f=M_f
  \quad\text{if the body elaborates to }t_f.
\end{array}
\]
For a method declaration owned by $S$, elaboration emits
$\mu(S,m)=M_{S,m}$, where $M_{S,m}$ is the same function wrapper as $M_f$ except that it has an additional explicit receiver parameter
$\kw{self}:\trans{S}$ before the ordinary method parameters.
An extension $S:P$ elaborates to a dictionary record
\[
  \delta(S,P)=\{\overline{m_i=\mu(S,m_i)}\}
\]
whose fields are the implementations required by $P$.
The surface conformance check ensures that this record has type $\Dict(P,\trans{S})$.

\section{Elaboration Safety}

This report proves safety of the surface-to-core elaboration, not progress and preservation for the core calculus itself.
We assume the standard core soundness theorem.
Let $\has_c$ denote the core typing judgment.
The proof uses the standard core typing rules for variables, functions, application, type abstraction and application, records, projection, conditionals, fixpoints, natural-number primitives, existential packing, and existential unpacking.
Typing of generated programs is phrased against a pre-collected translated global signature.
This matches the surface checker, which collects all top-level declarations before checking bodies.
Let $\mathcal{G}_\Psi$ contain top-level values $x:\trans{\Psi_V(x)}$, method functions $\mu(S,m):\trans{\Psi_M(S)(m)}$, and conformance dictionaries $\delta(S,P):\Dict(P,\trans{S})$ for every collected conformance $(S,P)$.
The declaration-preservation theorem below proves that the generated definitions assigned to these names actually inhabit the types in $\mathcal{G}_\Psi$.
Let
\[
  \mathcal{E}
  =
  \mathcal{G}_\Psi;\trans{\Omega};\trans{\Gamma};\trans{\Delta}
\]
be the translated core context.
Throughout the proof we assume the standard well-formedness invariants produced by surface checking:
$\Psi$ is well formed; $\Omega$ contains only declared protocol constraints; $\Gamma$ contains only well-formed surface types; $\Delta$ contains exactly the dictionary variables for the active constraints in $\Omega$; all struct fields, method names, and protocol requirements are distinct within their owners; and protocol requirement signatures are well formed under the protocol \Self resolver.
The theorem below proves preservation for successful elaborations.
The implementation performs elaboration after resolution and checking, so the elaboration relation is syntax-directed on checked ASTs; totality of that implementation pass is not part of the formal statement.

\subsection{Supporting Lemmas}

\paragraph{Lemma 1: Type translation is total and preserves well-formedness.}
If $\Psi;\Omega;\rho\has\tau\ok$, then $\trans{\tau}$ is defined and
\[
  \trans{\Psi};\trans{\Omega}\has_c\trans{\tau}\ok.
\]

\paragraph{Proof.}
By mutual induction with Lemma 2 on type well-formedness and protocol dictionary formation.
Base types translate by definition and are well formed by the corresponding core well-formedness rules.
For a struct type $S$, the premise $S\in\dom(\Psi_S)$ makes $\Psi_S(S)$ defined, so $\trans{S}$ is the record type obtained by translating each declared field type.
Those field types are well formed because program well-formedness checks declarations before elaboration; the induction hypothesis gives their translated field types, and the core record-type rule gives $\trans{S}\ok$.
For a type variable $X$, well-formedness gives $X\in\dom(\Omega)$, and $\trans{X}=X$ is well formed in $\trans{\Omega}$.
For \Self, well-formedness requires $\rho\neq\bot$; in a concrete receiver context \Self has already been resolved to the receiver struct, and in a protocol requirement it is translated only by the self-parameterized translation used in dictionary formation.
For function schemes, every source type parameter becomes a core type variable, every protocol constraint becomes a dictionary argument type, and ordinary parameter/result types are well formed by the induction hypothesis.
Dictionary argument types are well formed by Lemma 2.
For $\Any{P}$, the translated type is $\exists X.\{\kw{value}:X,\kw{dict}:\Dict(P,X)\}$.
Under fresh $X$, the field type $X$ is well formed and $\Dict(P,X)$ is well formed by Lemma 2, so the translation is defined and the existential type is well formed.
\qed

\paragraph{Lemma 2: Dictionary types are well formed.}
If $P\in\dom(\Psi_P)$ and $\trans{\Psi};\trans{\Omega}\has_c A\ok$, then
\[
  \trans{\Psi};\trans{\Omega}\has_c\Dict(P,A)\ok.
\]

\paragraph{Proof.}
This is the dictionary-formation component of the mutual induction used in Lemma 1.
Since $P$ is declared, every requirement scheme in $\Psi_P(P)$ is well formed under resolver $P$.
The self-parameterized translation $\transSelf{A}{\sigma_i}$ is well formed: it follows the ordinary translation for all constructors and translates each occurrence of protocol \Self as the well-formed core type $A$.
The dictionary record is therefore well formed.
\qed

\paragraph{Lemma 3: Context translation is well formed.}
If $\Psi;\Omega;\Gamma;\rho$ is a well-formed surface context, $\Delta$ maps only constraints present in $\Omega$, and $\mathcal{G}_\Psi$ is well formed, then $\mathcal{E}$ is a well-formed core context.

\paragraph{Proof.}
Each type variable in $\Omega$ becomes a core type variable.
Each local variable $x:\tau$ becomes $x:\trans{\tau}$, which is well formed by Lemma 1.
Each evidence binding $\Delta(X,P)=d$ becomes $d:\Dict(P,X)$, which is well formed by Lemma 2.
The translated global bindings are well formed by the premise on $\mathcal{G}_\Psi$.
\qed

\paragraph{Lemma 4: Translation commutes with type substitution.}
If $\tau'$ is well formed and $\tau$ is well formed under $X$, then
\[
  \trans{\subst{\tau'}{X}\tau}
  =
  \subst{\trans{\tau'}}{X}\trans{\tau}.
\]
The same equality holds pointwise for parameter lists, function schemes, and environments.

\paragraph{Proof.}
By structural induction on $\tau$.
Base types and nominal struct types contain no occurrence affected by the substitution.
The type-variable case is immediate: if the variable is $X$, both sides are $\trans{\tau'}$; otherwise both sides are the same variable.
For function schemes, apply the induction hypothesis to every parameter and result type, alpha-renaming bound type variables to avoid capture.
For $\Any{P}$, the protocol name is nominal and not affected by substitution.
The environment and scheme cases are pointwise extensions of the type case.
\qed

\paragraph{Lemma 5: Resolved \Self substitution matches self-parameterized translation.}
If a protocol requirement type $\tau$ is well formed under resolver $P$, and $S$ is a declared struct, then
\[
  \trans{[\Self\mapsto S]\tau}=\transSelf{\trans{S}}{\tau}.
\]
The same holds pointwise for parameter lists and function schemes.

\paragraph{Proof.}
By structural induction on $\tau$.
The \Self case is the only non-homomorphic case: the left side resolves \Self to $S$ and then translates it to $\trans{S}$, while the right side translates protocol \Self directly as the supplied core self type $\trans{S}$.
All other constructors follow from the induction hypothesis.
\qed

\paragraph{Lemma 6: \Self-free types are stable under opening existentials.}
If $\Self\notin\fv(\tau)$, then for any core type $A$,
\[
  \transSelf{A}{\tau}=\trans{\tau}.
\]
The same holds pointwise for parameter lists.

\paragraph{Proof.}
By structural induction on $\tau$.
Because \Self is not free in $\tau$, the substitution never changes a leaf of the type syntax.
All type constructors are preserved by translation, so both sides are syntactically equal.
\qed

\paragraph{Lemma 7: Argument elaboration preserves typing.}
If the deterministic argument matcher checks actual arguments against formal parameters,
\[
  \Psi;\Omega;\Gamma;\rho
    \has \overline{a}\Leftarrow\overline{\ell_i:\tau_i},
\]
and elaborates them as
\[
  \Psi;\Omega;\Gamma;\rho;\Delta
    \has \overline{a}\Leftarrow\overline{\ell_i:\tau_i}
    \elab \overline{t_i},
\]
then for each $i$,
\[
  \mathcal{E}\has_c t_i:\trans{\tau_i}.
\]

\paragraph{Proof.}
This lemma is proved simultaneously with Theorem 2, ordered by the size of the elaboration derivation.
Proceed by induction on the argument-matching derivation.
The empty case is immediate.
For a positional argument, the surface rule has a premise
$\Psi;\Omega;\Gamma;\rho\has e:\tau_i$, and elaboration has the corresponding premise
$\Psi;\Omega;\Gamma;\rho;\Delta\has e:\tau_i\elab t_i$.
The simultaneous induction hypothesis for expression elaboration, applied to this smaller derivation, gives
$t_i:\trans{\tau_i}$; the induction hypothesis handles the remaining arguments.
The labeled case is identical after the deterministic matcher selects the corresponding formal parameter and orders the typed AST.
\qed

\paragraph{Lemma 8: Function wrappers preserve typing.}
Let
\[
  \sigma=
  \ftype{\overline{X_i:\overline{P_i}}}
         {\overline{\ell_j:\tau_j}}
         {\tau_r}.
\]
If
\[
  \mathcal{G}_\Psi;
  \trans{\Omega,\overline{X_i:\overline{P_i}}};
  \trans{\Gamma,\overline{x_j:\tau_j}};
  \trans{\Delta,\overline{(X_i,P)=\delta(X_i,P)}}
  \has_c t_f:\trans{\tau_r},
\]
then the generated wrapper
\[
  \Lambda \overline{X_i}.\
  \lambda \overline{\delta(X_i,P):\Dict(P,X_i)}.\
  \lambdastar\ \overline{x_j:\trans{\tau_j}}.\ t_f
\]
has type $\trans{\sigma}$ in $\mathcal{E}$.
The same statement holds for method wrappers after adding the explicit receiver parameter $\kw{self}:\trans{S}$ as the first ordinary parameter.

\paragraph{Proof.}
By repeated use of the core rules for type abstraction and function abstraction.
Each source type parameter becomes a core type variable.
Each protocol constraint $X_i:P$ becomes a dictionary parameter of type $\Dict(P,X_i)$.
Each ordinary parameter $x_j:\tau_j$ becomes a core parameter $x_j:\trans{\tau_j}$.
The premise gives the body type $\trans{\tau_r}$ under exactly this extended core context.
The zero-argument case is covered by the $\lambdastar$ convention, which abstracts over a single \Unit\ argument and therefore inhabits $\Unit\to\trans{\tau_r}$.
The method case is identical, with the receiver parameter included in the ordinary parameter list.
\qed

\subsection{Preservation Theorems}

\paragraph{Theorem 1: Evidence elaboration preserves typing.}
Assume $\mathcal{G}_\Psi$ contains the translated global bindings prescribed above.
If
\[
  \Psi;\Omega\has\tau\modelsP P
  \quad\text{and}\quad
  \Psi;\Omega;\Delta\has\tau\modelsP P\elab d,
\]
then
\[
  \mathcal{G}_\Psi;\trans{\Omega};\trans{\Delta}
  \has_c d:\Dict(P,\trans{\tau}).
\]

\paragraph{Proof.}
By cases on evidence elaboration.
In \textsc{Ev-Struct}, $\tau=S$ and $(S,P)\in\dom(\Psi_C)$.
The translated global context contains the conformance dictionary binding
$\delta(S,P):\Dict(P,\trans{S})$ by the premise on $\mathcal{G}_\Psi$.
Since $\trans{\tau}=\trans{S}$, the desired type follows by the core variable rule.
In \textsc{Ev-TVar}, $\tau=X$, $X:\overline{P}\in\Omega$, and $\Delta(X,P)=d$.
Context translation puts $d:\Dict(P,X)$ in $\trans{\Delta}$, and $\trans{X}=X$.
\qed

\paragraph{Theorem 2: Expression elaboration preserves typing.}
If
\[
  \Psi;\Omega;\Gamma;\rho\has e:\tau
  \quad\text{and}\quad
  \Psi;\Omega;\Gamma;\rho;\Delta\has e:\tau\elab t,
\]
then
\[
  \mathcal{E}\has_c t:\trans{\tau}.
\]

\paragraph{Proof.}
By simultaneous induction on expression elaboration and argument elaboration derivations, using the corresponding surface typing rule.
Lemma 7 is the argument-elaboration component of this simultaneous induction.

In \textsc{E-Int}, the elaborated term is the same numeral $n$.
Core numerals have type $\Nat$, and $\trans{\Int}=\Nat$, so
$\mathcal{E}\has_c n:\trans{\Int}$.
The \textsc{E-Bool} and \textsc{E-String} cases are identical: the generated terms are the same literals, core typing gives $b:\Bool$ and $s:\String$, and the type translation maps $\Bool$ and $\String$ to those same core types.

In \textsc{E-Var}, either $x:\tau\in\Gamma$ or $\Psi_V(x)=\tau$.
By definition of context translation, $\mathcal{E}$ contains the corresponding core binding
$x:\trans{\tau}$, either in $\trans{\Gamma}$ or in $\mathcal{G}_\Psi$.
The core variable rule therefore gives
$\mathcal{E}\has_c x:\trans{\tau}$.
The \textsc{E-Self} case is the same argument specialized to the local receiver binding
$\kw{self}:S\in\Gamma$, giving
$\mathcal{E}\has_c \kw{self}:\trans{S}$.

For \textsc{E-Not}, the premise elaborates the operand as
$\ectx\has e:\Bool\elab t$.
The induction hypothesis gives $\mathcal{E}\has_c t:\Bool$.
The generated term is
$\kw{if}\ t\ \kw{then}\ \kw{false}\ \kw{else}\ \kw{true}$.
The guard has type $\Bool$, both branches have type $\Bool$, and the core conditional rule gives the required result type $\Bool=\trans{\Bool}$.
For \textsc{E-And}, the generated term is
$\kw{if}\ t_1\ \kw{then}\ t_2\ \kw{else}\ \kw{false}$.
The premises and induction hypotheses give $t_1:\Bool$ and $t_2:\Bool$; hence the conditional has type $\Bool$.
For \textsc{E-Or}, the generated term
$\kw{if}\ t_1\ \kw{then}\ \kw{true}\ \kw{else}\ t_2$
is typed by the same core conditional rule.

For \textsc{E-Add}, the premises elaborate the operands to $t_1$ and $t_2$ with surface type $\Int$.
The induction hypotheses give $t_1:\Nat$ and $t_2:\Nat$.
The closed operator abbreviation satisfies
$\has_c\mathsf{add}:\Nat\to\Nat\to\Nat$, so two applications yield
$\mathsf{add}\ t_1\ t_2:\Nat=\trans{\Int}$.
\textsc{E-Sub} is identical using
$\mathsf{sub}:\Nat\to\Nat\to\Nat$.
For \textsc{E-Lt}, the same operand types and
$\mathsf{lt}:\Nat\to\Nat\to\Bool$ give
$\mathsf{lt}\ t_1\ t_2:\Bool$.
\textsc{E-Gt} swaps the operands and still has type $\Bool$.
For \textsc{E-Le} and \textsc{E-Ge}, the inner $\mathsf{lt}$ application has type $\Bool$, and the generated conditional has Boolean branches, so the result type is $\Bool$.

For \textsc{E-Fun}, let
\[
  \sigma=
  \ftype{\overline{X_i:\overline{P_i}}}
         {\overline{\ell_j:\tau_j}}
         {\tau_r}.
\]
The elaboration premise checks the body under
$\Omega'=\Omega,\overline{X_i:\overline{P_i}}$,
$\Delta'=\Delta,\overline{(X_i,P)=\delta(X_i,P)}$,
and $\Gamma_f=\Gamma,\overline{x_j:\tau_j}$.
The induction hypothesis gives
$\mathcal{G}_\Psi;\trans{\Omega'};\trans{\Gamma_f};\trans{\Delta'}
\has_c t_f:\trans{\tau_r}$.
Lemmas 1 and 2 justify the well-formed translated parameter and dictionary types in this extended context.
Lemma 8 then types
\[
  \Lambda \overline{X_i}.\
  \lambda \overline{\delta(X_i,P):\Dict(P,X_i)}.\
  \lambdastar\ \overline{x_j:\trans{\tau_j}}.\ t_f
\]
at $\trans{\sigma}$, which is exactly the type translated from the surface function expression.

For \textsc{E-TypeApp}, the callee premise gives an elaboration
\[
  \ectx\has e:
  \ftype{\overline{X_i:\overline{P_i}}}
         {\overline{\ell_j:\tau_j}}
         {\tau_r}
  \elab t.
\]
The induction hypothesis gives $t$ the translated universal type.
Core type application instantiates each universal binder $X_i$ with
$\trans{\tau_i'}$.
For every source constraint $X_i:P$, the evidence premise elaborates to a dictionary $d_{i,P}$; by Theorem 1,
$d_{i,P}:\Dict(P,\trans{\tau_i'})$.
After the type applications and these dictionary applications, the remaining core type is
\[
  \arrstar(
    \overline{\trans{\subst{\overline{\tau_i'}}{\overline{X_i}}\tau_j}},
    \trans{\subst{\overline{\tau_i'}}{\overline{X_i}}\tau_r}).
  )
\]
By Lemma 4, this is the translation of the surface substitution result stated in the conclusion.

For \textsc{E-Call}, the callee elaborates to $t$ at type
$\ftype{\emptyctx}{\overline{\ell_i:\tau_i}}{\tau_r}$.
The induction hypothesis gives
$t:\arrstar(\overline{\trans{\tau_i}},\trans{\tau_r})$.
Lemma 7 gives the matched and elaborated actual arguments
$\overline{u_i}$ with types $\overline{\trans{\tau_i}}$.
Repeated use of the core application rule, or the zero-argument convention encoded by $\appstar$, gives
$\appstar(t,\overline{u_i}):\trans{\tau_r}$.

For \textsc{E-Construct}, the premises provide
$\Psi_S(S)=\{\overline{x_i:\tau_i}\}$ and elaborated constructor arguments
$\overline{t_i}$ matched against those fields.
Lemma 7 gives each $t_i:\trans{\tau_i}$.
The core record-introduction rule gives
$\{\overline{x_i=t_i}\}:\{\overline{x_i:\trans{\tau_i}}\}$.
By definition, this record type is $\trans{S}$, so the constructor term has the translated struct type.
For \textsc{E-Field}, the receiver premise gives
$\ectx\has e:S\elab t$ and $\Psi_S(S)(x)=\tau$.
The induction hypothesis yields $t:\trans{S}$.
Since $\trans{S}$ is the record type containing field $x:\trans{\tau}$, the core projection rule gives
$t.x:\trans{\tau}$.

For \textsc{E-StructMethod}, write
\[
  \sigma=
  \ftype{\overline{X_i:\overline{P_i}}}
         {\epsilon:S,\overline{\ell_j:\tau_j}}
         {\tau_r}.
\]
The translated global context contains
$\mu(S,m):\trans{\sigma}$.
The receiver premise and induction hypothesis give $t:\trans{S}$.
In the generated wrapper $M_S$, the outer core type abstractions introduce the method type variables $X_i$.
The following dictionary lambdas introduce one parameter
$\delta(X_i,P):\Dict(P,X_i)$ for each method constraint.
Inside the value-lambda body, $\mu(S,m)$ is first type-applied to the current $X_i$, then applied to the introduced dictionaries, then applied to the captured receiver $t:\trans{S}$, and finally applied to the visible arguments $\overline{y_j:\trans{\tau_j}}$.
Core application therefore gives the body type $\trans{\tau_r}$.
Core abstraction over the visible arguments, dictionaries, and type variables reconstructs
$\trans{\ftype{\overline{X_i:\overline{P_i}}}{\overline{\ell_j:\tau_j}}{\tau_r}}$.

For \textsc{E-ConstrainedMethod}, the receiver has type variable $X$ and elaborates to $t$.
The induction hypothesis gives $t:X$.
The premise $\Delta(X,P_k)=d$ and context translation give
$d:\Dict(P_k,X)$.
If
\[
  \Psi_P(P_k)(m)=
  \ftype{\overline{Y_i:\overline{Q_i}}}
         {\epsilon:\Self,\overline{\ell_j:\tau_j}}
         {\tau_r},
\]
then dictionary projection gives $d.m$ the requirement type translated with protocol \Self interpreted as $X$.
The generated wrapper $M_X$ type-abstracts over $\overline{Y_i}$, abstracts over dictionaries for $\overline{Y_i:\overline{Q_i}}$, and abstracts over visible arguments
$\overline{y_j:\trans{[\Self\mapsto X]\tau_j}}$.
Its body applies $d.m$ to those type arguments, dictionaries, the receiver $t:X$, and the visible arguments.
This body has type $\transSelf{X}{\tau_r}$.
By Lemma 5, the parameter and result types are equal to the ordinary translations of the surface types after substituting $[\Self\mapsto X]$, so $M_X$ has the type stated by the elaboration rule.

For \textsc{E-IfElse}, the premises elaborate the guard and both branches:
$e\elab t$, $ss_1\elab t_1$, and $ss_2\elab t_2$.
The induction hypothesis gives $t:\Bool$ and, using Theorem 3 for statement sequences, the branch typings
$t_1:\trans{\tau}$ and $t_2:\trans{\tau}$.
The core conditional rule gives
$\kw{if}\ t\ \kw{then}\ t_1\ \kw{else}\ t_2:\trans{\tau}$.

For \textsc{E-AsAny}, the induction hypothesis gives $t:\trans{\tau}$.
The evidence premise and Theorem 1 give $d:\Dict(P,\trans{\tau})$.
The core record rule gives
\[
  \{\kw{value}=t,\kw{dict}=d\}:
  \{\kw{value}:\trans{\tau},\kw{dict}:\Dict(P,\trans{\tau})\}.
\]
Core existential introduction packages this record at witness type $\trans{\tau}$, producing type
$\exists X.\{\kw{value}:X,\kw{dict}:\Dict(P,X)\}$.
This is exactly $\trans{\Any{P}}$.

For \textsc{E-ExistentialMethod}, the induction hypothesis gives
\[
  t:\trans{\Any{P}}
   =\exists X.\{\kw{value}:X,\kw{dict}:\Dict(P,X)\}.
\]
The generated term first abstracts over method type variables, dictionary parameters, and visible arguments.
Inside those abstractions, the unpack rule introduces a fresh type $X$ and a package body variable
$z:\{\kw{value}:X,\kw{dict}:\Dict(P,X)\}$.
Core projection gives $z.\kw{value}:X$ and
$z.\kw{dict}:\Dict(P,X)$.
If the protocol requirement for $m$ is
\[
  \ftype{\overline{Y_i:\overline{Q_i}}}
         {\epsilon:\Self,\overline{\ell_j:\tau_j}}
         {\tau_r},
\]
then $z.\kw{dict}.m$ has this requirement type translated with protocol \Self interpreted as the opened type $X$.
After applying it to the method type arguments, the introduced requirement dictionaries, the hidden receiver $z.\kw{value}$, and the visible arguments, the unpack body has result type $\transSelf{X}{\tau_r}$.
The side conditions require
$\Self\notin\fv(\overline{\tau_j})$ and $\Self\notin\fv(\tau_r)$.
By Lemma 6, the visible parameter types and result type are therefore independent of the locally opened $X$:
$\transSelf{X}{\overline{\tau_j}}=\trans{\overline{\tau_j}}$ and
$\transSelf{X}{\tau_r}=\trans{\tau_r}$.
Thus the result type may escape the unpack, satisfying the core existential-elimination side condition, and the whole generated wrapper has the receiver-stripped method type translated from the surface rule.
\qed

\paragraph{Theorem 3: Statement-sequence elaboration preserves typing.}
If
\[
  \Psi;\Omega;\Gamma;\rho\has ss:\tau
  \quad\text{and}\quad
  \Psi;\Omega;\Gamma;\rho;\Delta\has ss:\tau\elab t,
\]
then
\[
  \mathcal{E}\has_c t:\trans{\tau}.
\]

\paragraph{Proof.}
By induction on the statement-sequence elaboration derivation.
\textsc{ES-Empty} follows from the core unit rule.
\textsc{ES-Expr-Last} follows directly from Theorem 2.
In \textsc{ES-Expr-Cons}, Theorem 2 gives $t_1:\trans{\tau_1}$, and the induction hypothesis gives $t_2:\trans{\tau_2}$.
The generated term $(\lambda\_:\trans{\tau_1}.\ t_2)\ t_1$ has type $\trans{\tau_2}$ by core abstraction and application.
In \textsc{ES-Let}, Theorem 2 gives the initializer $t_x:\trans{\tau_x}$.
The induction hypothesis, applied under the extended context $\Gamma,x:\tau_x$, gives the body $t:\trans{\tau}$.
The generated lambda application therefore has type $\trans{\tau}$.
For a local function declaration, the induction hypothesis gives the body of the local function in the context extended with its type parameters, dictionaries, and value parameters.
Lemma 8 types the generated function wrapper at $\trans{\sigma}$.
The induction hypothesis then types the remaining sequence under the extended binding for $f$, and the final lambda application preserves the result type.
\qed

\paragraph{Theorem 4: Declaration elaboration preserves typing.}
Let $\Psi$ be the pre-collected global environment for a well-formed program, and let $\mathcal{G}_\Psi$ be its translated global signature.
If a well-formed surface declaration elaborates to core definitions $C$, then every generated definition in $C$ type checks against $\mathcal{G}_\Psi$ at the type assigned to its name by $\mathcal{G}_\Psi$.
In particular, top-level value, function, method, and conformance dictionary definitions inhabit the types prescribed by the translated global signature.

\paragraph{Proof.}
By cases on declaration elaboration.
Struct and protocol declarations emit no core term bindings; they only contribute entries to $\Psi$, whose type-level well-formedness follows from declaration checking and Lemmas 1 and 2.
For a top-level let declaration, surface typing gives $\Psi;\emptyctx;\emptyctx;\bot\has e:\tau$.
Expression preservation gives the elaborated initializer type $\trans{\tau}$, so the generated binding $x=t$ is well typed.

For a top-level function, statement-sequence preservation applies to the body under the function's type parameters, dictionary evidence parameters, and ordinary parameters.
Lemma 8 then gives the generated binding the translated function-scheme type.
For a method declaration owned by $S$, the same argument applies, except that the local context also contains the explicit receiver $\kw{self}:S$.
The resulting top-level method function has type $\trans{\Psi_M(S)(m)}$.

For an extension $S:P$, the surface conformance check ensures that each requirement
$m_i:\sigma_i$ of $P$ is implemented by a method of $S$ whose type is
$[\Self\mapsto S]\sigma_i$.
The method case above gives
$\mu(S,m_i):\trans{[\Self\mapsto S]\sigma_i}$.
By Lemma 5,
$\trans{[\Self\mapsto S]\sigma_i}=\transSelf{\trans{S}}{\sigma_i}$.
This is exactly the field type expected by $\Dict(P,\trans{S})$.
The core record-introduction rule therefore gives
$\delta(S,P)=\{\overline{m_i=\mu(S,m_i)}\}:\Dict(P,\trans{S})$.
\qed

\paragraph{Theorem 5: Program elaboration preserves typing.}
If $\has p\ok$ and $p\elab I$, then the generated core program $I$ is well typed.

\paragraph{Proof.}
Program checking first collects a global environment $\Psi$ and then checks each declaration against it.
Elaboration emits core definitions for the names in the pre-collected translated signature $\mathcal{G}_\Psi$.
Apply Theorem 4 to every emitted declaration definition.
The result is a well-typed core program whose global signature contains all top-level values, functions, methods, and conformance dictionaries required by expression and evidence elaboration.
For every elaborated expression or statement sequence inside the program, Theorems 2 and 3 give the corresponding translated core type.
Thus the generated core program is well typed.
\qed

\paragraph{Corollary: End-to-end safety.}
If $\has p\ok$, $p\elab I$, and the standard core calculus is type sound, then executing $I$ cannot get stuck.
This follows directly from Theorem 5 and the assumed core progress and preservation theorems.

\section{Implementation Notes}

The implementation follows the same staged structure as the formal development, but it is intentionally small and direct.
The executable entry point in \texttt{src/interpreter.mbt} reads one source file, tokenizes it with the surface lexer, parses it with the generated surface parser, resolves names, checks types, elaborates the typed surface program to core concrete syntax, and finally invokes the existing core parser, typechecker, and evaluator.
The debugging entry point in \texttt{src/surface/parser/main.mbt} runs the same front-end pipeline but prints the typed surface program and generated core program instead of executing the core.

The front end is split into four packages under \texttt{src/surface/parser}.
The \texttt{syntax} package contains the token definitions, grammar, source-position helpers, and the first AST.
The \texttt{resolver} package performs the first semantic pass:
it collects top-level struct and protocol names, resolves nominal type references and type variables, inserts explicit receiver parameters for methods, and records whether \Self is being resolved in a struct, protocol, extension, or top-level context.
The \texttt{checker} package builds a global environment, checks protocol conformances, resolves fields and methods, matches Swift-style argument labels, enforces generic protocol constraints at type application, rejects unsafe existential method selection, and produces a typed AST.
The \texttt{elaborator} package then performs dictionary-passing translation from that typed AST to core syntax.

The elaborator first collects enough global information to render struct record types, protocol dictionary types, method functions, and extension dictionaries.
Surface structs become core records.
Protocol conformances become dictionary records named by the implementing struct and protocol.
Generic constraints become explicit dictionary parameters.
Method definitions become top-level core functions with an explicit receiver parameter.
Method selection returns a receiver-capturing wrapper, so generic method values can be selected before their type arguments are supplied.
Existential values are rendered as core packages containing a hidden representation type, a value, and a dictionary; existential-safe method calls are rendered by opening the package and immediately calling the dictionary method on the hidden receiver.
Surface arithmetic and Boolean operators are not primitive core syntax; they are rendered as small core terms built from conditionals, fixpoints, and natural-number primitives.

\section{Implementation Status and Limitations}

The implementation already supports the proof-relevant path described above:
Swift-style parameter labels, structs, protocols, extensions, constrained generic functions, dictionary-passing elaboration, existential packaging with $e\ \kw{as}\ \Any{P}$, and existential-safe method calls, including method-local type parameters.
The arithmetic and boolean operators are surface syntax that elaborates into existing core terms.

Important limitations remain.
There is no surface-level explicit unpack.
Surface $\Int$ is currently represented by core $\Nat$, so it does not match Swift's signed integer semantics.
String concatenation is not implemented because the core language lacks the primitive.
The report also treats the core calculus's soundness theorem as an assumption rather than reproving it.

\end{document}
